Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-07-07 POPL/paper.tex
more paper organization

This result feels similar in spirit to the definition of action
algebras, which are algebras where the product also has adjoint
operations which results in the Kleene star being more easily
axiomatized \cite{kozen1994}.

@@

\begin{lemma}[\cite{day1970}]
  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
  monoidal closed.
\end{lemma}

@@

\begin{definition}
  A \emph{Chomsky Category} is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

@@

\steven{TODO: move to related work section at the end}

@@

\subsection{Syntax}
\label{subsec:syntax}

In this section we will walk through most of the syntax of our type theory. We
include the full syntax in \cref{sec:infer}. \steven{full syntax in appendix}

The linear typing judgment in our syntax takes on the following schematic form,
\[
  \Gamma ; \Delta \vdash M : A
\]

Let us briefly digest what this judgment is trying to say. First, $A$ represents
a \emph{linear type} in our syntax. The intended semantics of these linear types
are formal grammars. That is, the linear typing system is designed to syntactically
reflect the behavior of formal grammars. For this reason, we may often
interchangeably use the terms ``linear type'' and ``grammar''. To see the
correspondence between types and grammars, note that the constructors for linear types
mirror operations on formal grammars. For example, we have linear type constructors
$\otimes$, $\oplus$, and $\amp$ which correspond to grammar concatenation,
disjunction, and conjunction, respectively. The other type constructors also have
corresponding grammar intuition which may be seen in \cref{subsubsec:other}.

$M$ is an inhabitant of type $A$, which we want to think of as a parse tree of
the grammar $A$. The core idea of this entire paper follows precisely from this
single correspondence: grammars are types, and the inhabitants of these types
are parse trees for the grammars!

$\Gamma$ represents a \emph{non-linear context} --- also referred to as an
\emph{intuitionistic} context. The non-linear fragment of the theory
corresponds to usual Martin-L\"of type theory. That is, the intuitionistic
contexts obey all of the usual structural rules --- contraction, weakening, and
exchange --- so we may use their contents as many times as we'd like.

On the other hand $\Delta$ represents a \emph{linear context}. These linear
contexts
behave substructurally --- in particular they have no contraction, weakening, or exchange
rule. This means that everything within $\Delta$ must be used \emph{exactly
  once} and \emph{in order of occurrence}. Thus, we can think of the linear
contexts as an ordered list of limited
resources. Once a resource is consumed, we cannot make reference to it again.
We may then
think of the variables in a linear context akin to building blocks for
constructing patterns over strings.

@@

\todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
\max{this is a good point to give an overview of how regular expressions and context-free grammars/expressions are essentially sub-languages of our type system.}

@@

\subsection{Semantics in Languages}
Recall from \cref{sec:synindsyn} that every formal grammar induces a formal language. This corresponds to another
interpretation $L(\cdot)$ of our syntax. In particular, the semantics of a
linear type $A : L_{i}$
from \cref{subsec:seming} are augmented such that

@@

\todoin{Derivation tree for this construction}

@@

\appendix

\section{Inference Rules}
\label{sec:infer}

\begin{figure}[h!]
  \label{fig:structjdg}
  \begin{mathpar}
    \inferrule{~}{\ctxwff \cdot}
    \and
    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}

  \end{mathpar}
  \caption{Structural judgments}
\end{figure}

\begin{figure}
  \label{fig:typewf}
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
 %
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
%
    \and
%
    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash 1 : U_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash I : L_i}
 %
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
    %
    \and
    %
    \inferrule{~}{\Gamma \vdash \top : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
%
    \\
    %
    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
    %
    \and
    %
    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
  \end{mathpar}
  \caption{Type well-formedness}
\end{figure}

\begin{figure}
  \label{fig:jdgeq}
  \begin{mathpar}
    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
%
    \and
%
    \inferrule{~}{}
%
    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
%
    \and
%
    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
%
    \and
%
    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
\end{mathpar}
  \caption{Judgmental equality}
\end{figure}

>>>>

This result feels similar in spirit to the definition of action
algebras, which are algebras where the product also has adjoint
operations which results in the Kleene star being more easily
axiomatized \steven{TODO cite}
% \cite{kozen1994}.

@@

\begin{lemma}[TODO cite
  % \cite{day1970}
  ]
  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
  monoidal closed.
\end{lemma}

@@

\begin{definition}
  A \emph{Chomsky category} is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

@@

@@

The structrual judgements are shown in \Cref{fig:structjdg}, the typing well-formedness in \Cref{fig:typewf}, and the intuitionistic typing rules in \Cref{fig:inttyping}. There are all just as they appear in \cite{krishnaswami_integrating_2015}.

Our contribution is the addition of base types drawn from the alphabet $\Sigma$ along with the omission of the exchange rule for linear typing judgements.

\steven{TODO need to update syntax in these rules}

\begin{figure}
  \begin{mathpar}
    \boxed{\ctxwff \Gamma}

    \inferrule{~}{\ctxwff \cdot}
    \and
    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}

    \boxed{\linctxwff \Gamma \Delta}

    \boxed{\ctxwffjdg \Gamma X}

    \boxed{\linctxwffjdg \Gamma A}

    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}

    \boxed{\linctxwffjdg \Gamma {A \equiv B}}

  \end{mathpar}
  \caption{Structural judgments}
  \label{fig:structjdg}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
 %
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
%
    \and
%
    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash 1 : U_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash I : L_i}
 %
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
    %
    \and
    %
    \inferrule{~}{\Gamma \vdash \top : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
%
    \\
    %
    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
    %
    \and
    %
    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
  \end{mathpar}
  \caption{Type well-formedness}
  \label{fig:typewf}
\end{figure}

\begin{figure}
  \begin{mathpar}
  \boxed{\Gamma \vdash x : X}

  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
  %
  \\\
  %
  \inferrule{~}{\Gamma \vdash () : 1}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
  %
  \\
%
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
  \and
  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
  %
  \\
  %
  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
  \and
  \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
  \end{mathpar}
  \caption{Intuitionistic typing}
  \label{fig:inttyping}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \boxed{\Gamma ; \Delta \vdash a : A}

    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : I \\ \Gamma ; \Delta_1',\Delta_2' \vdash e' : C}{\Gamma ; \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \otimes B \\ \Gamma ; \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma ;  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta', \Delta \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta, \Delta' \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}{\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma ; \Delta \vdash \app e {e'} : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : X \quad \Gamma ; \Delta \vdash e' : \subst A e x}{\Gamma ; \Delta \vdash (e, e') : \LinSigTy x X A}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X ; \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma; \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \Delta \vdash () : \top}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e_1 : A_1 \quad \Gamma ; \Delta \vdash e_2 : A_2}{\Gamma ; \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : A_1 \amp A_2 }{\Gamma ; \Delta \vdash \pi_i \, e : A_i}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma ; \cdot \vdash \mathsf{G}^{-1}\, e : A}
    %
    \\
    %
    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
    \and
    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
  \end{mathpar}
  \caption{Linear typing}
  \label{fig:linsyntax}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \boxed{\Gamma \vdash e \equiv e' : X}

    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
%
    \and
%
    \inferrule{~}{}
%
    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
%
    \and
%
    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
%
    \and

    \boxed{\Gamma ; \Delta \vdash a \equiv a' : A}

    \\
%
    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
\end{mathpar}
  \caption{Judgmental equality}
  \label{fig:jdgeq}
\end{figure}
\subsection{Syntax}
\label{subsec:syntax}

\paragraph{Non-linear Types}

The non-linear fragment of the theory
corresponds to usual Martin-L\"of type theory. That is, the non-linear
contexts obey all of the usual structural rules --- contraction, weakening, and
exchange --- so we may use their contents as many times as we'd like. For the application of parsing, we think of these as primarily as indicies terms for indexed grammars. That is, they are as useful for their dependence. We may think of them as brining into scope definitions from ordinary mathematics to be used for indexing

\paragraph{Linear Types}

The linear typing judgment in our syntax takes on the following schematic form $\Gamma ; \Delta \vdash a : A$. First, $A$ represents
a \emph{linear type} in our syntax. The intended semantics of these linear types
are formal grammars. That is, the linear typing system is designed to syntactically
reflect the behavior of formal grammars. For this reason, we may often
interchangeably use the terms ``linear type'' and ``grammar''.

The term $a$ is an inhabitant of type $A$, which we want to think of as a parse tree of
the grammar $A$. The core idea of this entire paper follows precisely from this
single correspondence: grammars are types, and the inhabitants of these types
are parse trees for the grammars.

$\Gamma$ reprsents a non-linear context, while $\Delta$ represents a \emph{linear context}. These linear
contexts
behave substructurally --- in particular they have no contraction, weakening, or exchange
rule. This means that everything within $\Delta$ must be used \emph{exactly
  once} and \emph{in order of occurrence}. Thus, we can think of the linear
contexts as an ordered list of limited
resources. Once a resource is consumed, we cannot make reference to it again.
We may then
think of the variables in a linear context akin to building blocks for
constructing patterns over strings.

@@

% \todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
\max{this is a good point to give an overview of how regular expressions and context-free grammars/expressions are essentially sub-languages of our type system.}

@@

\subsection{Canonicity}

Let $\cdot \vdash A$ be a closed linear type. Then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $A$:
\begin{enumerate}
\item On the one hand we have the set-theoretic semantics just
  defined, $\llbracket A \rrbracket \cdot w$
\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
  \lceil w \rceil \vdash e : G$.
\end{enumerate}
It is not difficult to see that at least for the ``purely positive''
formulae (those featuring only the positive connectives
$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
that the nodes of the tree correspond precisely to the introduction
forms of the type. However it is far less obvious that \emph{every}
linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
sequence of introduction forms since proofs can include elimination
forms as well. To show that this is indeed the case we give a
\emph{canonicity} result for the calculus: that the parses for .

\begin{definition}
  A non-linear type $X$ is purely positive if it is built up using
  only finite sums, finite products and least fixed points.

  A linear type is purely positive if it is built up using only finite
  sums, tensor products, generators $c$, least fixed points and linear
  sigma types over purely positive non-linear types.
\end{definition}

\begin{definition}
  %% Let $X$ be a closed non-linear type. The closed elements $\textrm{Cl}(X)$ of $X$ are the definitional equivalence classes of terms $\cdot \vdash e : X$.

  Let $A$ be a closed linear type. The nerve $N(A)$ is a presheaf on
  strings that takes a string $w$ to the definitional equivalence
  classes of terms $\cdot; \lceil w\rceil \vdash e: N(A)$.
\end{definition}

\begin{theorem}[Canonicity]
  Let $A$ be a closed, purely positive linear type. Then there is an
  isomorphism between $\llbracket A\rrbracket$ and $N(A)$.
\end{theorem}
\begin{proof}
  We outline the proof here, more details are in the appendix. The
  proof proceeds first by a standard logical families construction
  that combines canonicity arguments for dependent type theory
  \cite{coquand,etc} with logical relations constructions for linear
  types \cite{hylandschalk}. It is easy to see by induction that the
  logical family for $A$, $\hat A$ is isomorphic to $\llbracket A
  \rrbracket$ and the fundamental lemma proves that the projection
  morphism $p : \hat A \to N(A)$ has a section, the canonicalization
  procedure. Then we establish again by induction that
  canonicalization is also a retraction by observing that introduction
  forms are taken to constructors.
\end{proof}

\begin{enumerate}
\item Every term $\lceil w \rceil \vdash p : G + H$ is equal to $\sigma_1q$ or $\sigma_2 r$ (but not both)
\item There are no terms $\lceil w \rceil \vdash p : 0$
\item If there is a term $\lceil w \rceil \vdash p : c$ then $w = c$ and $p = x$.
\item Every term $\lceil w \rceil \vdash p : G \otimes H$ is equal to $(q,r)$ for some $q,r$
\item Every term $\lceil w \rceil \vdash p : \epsilon$ is equal to $()$
\item Every term $\lceil w \rceil \vdash p : c$ is equal to $x:c$
\item Every term $\lceil w \rceil \vdash p : \mu X. G$ is equal to $\textrm{roll}(q)$ where $q : G(\mu X.G/X)$
\item Every term $\lceil w \rceil \vdash p : (x:A) \times G$ is equal
  to $(M,q)$ where $\cdot \vdash M : A$
\end{enumerate}

To prove this result we will use a logical families model. We give a
brief overview of this model concretely:
\begin{enumerate}
\item A context $\Gamma$ denotes a family of sets indexed by closing substitutions $\hat\Gamma : (\cdot \vdash \Gamma) \Rightarrow \Set_i$
\item A type $\Gamma \vdash X : U_i$ denotes a family of sets $\hat X : \Pi(\gamma:\cdot \vdash \Gamma) \hat\Gamma \Rightarrow (\cdot \vdash \simulsubst X \gamma) \Rightarrow \Set_i$
\item A term $\Gamma \vdash e : X$ denotes a section $\hat e : \Pi(\gamma)\Pi(\hat\gamma)\hat X \gamma \hat\gamma (\simulsubst e \gamma)$
\item A linear type $\Gamma \vdash A : L_i$ denotes a family of grammars $\hat A : \Pi(\gamma:\cdot\vdash\Gamma)\,\hat\Gamma \Rightarrow \Pi(w:\Sigma^*) (\cdot;\lceil w\rceil \vdash A[\gamma])\Rightarrow \Set_i$, and the denotation of a linear context $\Delta$ is similar.
\item A linear term $\Gamma;\Delta \vdash e : A$ denotes a function \[\hat e : \Pi(\gamma)\Pi(\hat\gamma)\Pi(w)\Pi(\delta : \lceil w \rceil \vdash \simulsubst \Delta \gamma) \hat\Delta \gamma \hat\gamma \delta \Rightarrow \hat A \gamma \hat\gamma w {(\simulsubst {\simulsubst e \gamma} \delta)}\]
\end{enumerate}
And some of the constructions:
\begin{enumerate}
\item $\widehat {(G A)} \gamma \hat\gamma e = \hat A \gamma \hat\gamma \varepsilon (G^{-1}e)$
\item $\widehat {(A \otimes B)} \gamma \hat\gamma w e = \Sigma(w_Aw_B = w)\Sigma(e_A)\Sigma(e_B) (e_A,e_B) = e \wedge \hat A \gamma \hat \gamma w_A e_A \times \hat B \gamma \hat \gamma w_B e_B$
\item $\widehat {(A \lto B)} \gamma \hat\gamma w e = \Pi(w_A)\Pi(e_A) \hat A \gamma\hat\gamma w_A e_A \Rightarrow \hat B \gamma\hat\gamma (ww_A) (\applto {e_A} e)$
\end{enumerate}

First, the category with families will be
the category of logical families over set contexts/types
$\Delta$/$A$. Then the propositional portion will be defined by
mapping a logical family $\hat \Gamma \to \Gamma$

First, let $L$ be the category of BI formulae and proofs (quotiented
by $\beta\eta$ equality). Define a functor $N : L \to \Set^{\Sigma^*}$ by
\[ N(\phi)(w) = L(w,\phi) \]

Then define the gluing category $\mathcal G$ as the comma category
$\Set^{\Sigma^*}/N$. That is, an object of this category is a pair of
a formula $\phi \in L$ and an object $S \in \mathcal
P(\Sigma^*)/N(\phi)$. We can then use the equivalence $\mathcal
P(\Sigma^*)/N(\phi) \cong \mathcal P(\int N(\phi))$ to get a simple
description of such an $S$: it is simply a family of sets indexed by
proofs $L(w,\phi)$:
\[ \prod_{w\in\Sigma^*} L(w,\phi) \to \Set \]
This category clearly comes with a projection functor $\pi : \mathcal
G \to \mathcal L$ and then our goal is to define a section by using
the universal property of $\mathcal L$.

To this end we define
\begin{enumerate}
\item $(\phi, S) \otimes (\psi, T) = (\phi \otimes \psi, S\otimes T)$ where
  \[ (S \otimes T)(w, p) = (w_1w_2 = w) \times (q_1,q_2 = p) \times S\,w_1\,q_1 \times T\,w_2\,q_2\]
\item $(\phi, S) \multimap (\psi, T) = (\phi \multimap \psi, S \multimap T)$ where
  \[ (S \multimap T)(w,p) = w' \to q \to S\,w'\,q \to T (ww') (p\,q) \]
\item $\mu X. ??$ ??
\end{enumerate}

\subsection{Semantics in Languages}
% Recall from \cref{sec:synindsyn} that every formal grammar induces a formal language. This corresponds to another
interpretation $L(\cdot)$ of our syntax. In particular, the semantics of a
linear type $A : L_{i}$
from \cref{subsec:seming} are augmented such that

@@

% \todoin{Derivation tree for this construction}

@@

\end{document}
